---
title: "Intermediate R - Control Structures"
output:
  html_document:
    df_print: paged
---

This lesson is about control structures which are statements or commands that will alter the sequential execution of statements.

The first control structure/statement is the if statement and its variants. We will use the iris dataset.

```{r}
x <- iris$Petal.Length
n <- length(x)
if (x[n] > 2) {
  print("1st element is medium")
  print(iris$Species[1])
}
```
Next is the if else statement.

```{r}
if (x[1] <= 2) {
  print("short")
} else {
  print("long")
}
```
Next is the if else if.

```{r}
if (x[1] >= 2) {
  print("short")
} else if (x[1] <= 0.5) {
  print("medium")
  iris$Species[1]
} else if (x[1] < 1) print("between 0.5 and 1") else print("between 1 and 2")
```
The next control structure is the for loop.

This code will display the word "hello" 10 times.

```{r}
for (i in 1:10) {
  print("hello")
}
```
This code will display the word "hello" 5 times.
```{r}
j <- 2:6
for (i in j) {
  print("hello")
}
```
This code will display the 1st 10 petal lengths.

```{r}
for (i in 1:10) {
  print(x[i])
}
```
```{r}
x.inches <- numeric()
for (i in 1:10)
{
  x.inches[i] <- x[i] / 2.54
  print(x[i])
}
x.inches
```
The following R chunk will create a new vector named category which will classify the first 10 flowers in the iris data set as either short if its petal length is less than 1 cm or medium if its petal length is between 1 and 2 cm or long if its petal length is greater than 2 cm.
```{r}
category <- character()
x <- iris$Petal.Length
for (i in 1:10) {
  if (x[i] < 1) category[i] <- "short" else if (x[i] >= 1 & x[i] <= 2) category[i] <- "medium" else category[i] <- "long"
}
category
```

Modify the code so it will do the classification for all flowers in the iris data set.

```{r}
category <- character()
x <- iris$Petal.Length
n <- length(x)
for (i in 1:n) {
  if (x[i] < 1) {
    category[i] <- "short"
  } else if (x[i] >= 1 & x[i] <= 2) {
    category[i] <- "medium"
  } else {
    category[i] <- "long"
  }
}
category
```

What percent of the iris data set are classified as short, medium and long?

```{r}
table(category) / n * 100
```

The case_when command within the dplyr can also be used to create the category variable.

```{r}
library(dplyr)
new_iris <- iris %>% mutate(category = case_when(Petal.Length < 1 ~ "short", Petal.Length >= 1 & Petal.Length <= 2 ~ "medium", TRUE ~ "long"))
new_iris$category
```

