---
title: "Functions in R"
output:
  html_document:
    df_print: paged
---
R has many built-in functions such as the mean(), median(), sd(), quantile(), etc.  All functions require an input called its argument and after calling the function will display an output.  As an example, the mean() function will display the mean of its input.

```{r}
x <- iris$Petal.Length
mean(x)
```

We can create user-defined functions in R using the function command. As an example, let's create a new function named average that will compute the mean of an input x.

```{r}
average <- function(x)
{n<-length(x)
 val <- sum(x)/n
 return(val)}
```
To check that the new function average() is working, let's try it with the iris$Petal.Length data and verify its computing the correct mean with the mean() function.
```{r}
average(iris$Petal.Length)
mean(iris$Petal.Length)
```
As another example, let's create a new function named zscore that will compute the zscores of the elements of a numeric input y.
```{r}
zscore <- function(y)
{out <- (y-mean(y))/sd(y)
 return(out)}
```

To test this function, we can again use the iris$Petal.Length as the input to this function when we call the function.
```{r}
zscore(iris$Petal.Length)
```
A function can have more than one input. As an example, let's create a function that will display the smaller mean of 2 vectors x and y.

```{r}
Min_mean <- function(x,y)
{x1<-mean(x)
x2<-mean(y)
 if (x1 < x2) return(list(x1,"first input has the smaller mean")) else return(list(x2,"2nd input has the smaller mean"))}

```

To test this function, we will call this function using the Petal.Length and Petal.Width from the iris data set as input.

```{r}
Min_mean(iris$Petal.Length,iris$Petal.Width)
```
As an exercise, create a function that will convert measurements from cm to inches or vice-versa depending on the unit of measurement of the data.  Name the function convert with input x and unit (a non-numeric object).  Test this function using the iris$Petal.Length data which are in "cm".  

Here's the R chunk that will create this function:
```{r}
convert <- function(x,unit)
{if (unit =="cm") {out <- x/2.54
                  return(list(out,"inches"))} else if                        (unit == "inches"){out <- x*2.54
                      return(list(out,"cm"))} 
}
```

This will test the function: 
```{r}
convert(iris$Petal.Length,"cm")
```
