---
title: "Models with 2 Predictors"
output:
  html_document:
    df_print: paged
---

Let's fit a no interaction model for predicting Petal.Length based on Petal.Width and Sepal.Length.

```{r}
no_interaction<- lm(Petal.Length~Petal.Width + Sepal.Length, data = iris)
summary(no_interaction)
```
The estimate of Petal.Width coefficient is 1.74810 so for every 1 cm increase in Petal.Width, Petal.Length is expected to increase by about 1.75 cm for iris flowers of the same Sepal.Length.

The estimate of Sepal.Length coefficient is 0.54226 so for every 1 cm increase in Sepal.Length, Petal.Length is expected to increase by about 0.54 cm for iris flowers of the same Petal.Width.

Let's now fit an interaction model for predicting Petal.Length based on Petal.Width and Sepal.Length.

```{r}
interaction<- lm(Petal.Length~Petal.Width * Sepal.Length, data = iris)
summary(interaction)
```

Note the use of * instead of + when the predictors are specified in an interaction model and note the additional interaction coefficient in the result.

In this model, the estimate of Petal.Width coefficient of 2.97115 is no longer the rate of change of Petal.Length per 1 cm increase in Petal.Width for iris flowers of the same Sepal.Length.  The rate of change of Petal.Length per 1 cm increase in Petal.Width is now 2.97115 + -0.22248 * Sepal.Length for iris flowers with that Sepal.Length.

Note that the improvement of this interaction model compared with the no interaction model is marginal so this is a situation where the simpler model is preferable.

Let us now fit a parallel slopes model using Petal.Width and Species as predictors since both have a realtionship with Petal.Length as indicated by the scatterplot and correlation between the 2 numeric variables and by the side-by-side boxplots below for the numeric and non-numeric variables.

```{r}
library(ggplot2)
ggplot(iris,aes(x=factor(Species),y=Petal.Length))+geom_boxplot()
```
It's clear that knowing the specie is useful in predicting the Petal.Length so we will include Specie as a predictor in the model.

Let's fit a parallel slopes model for predicting Petal.Length using Petal.Width and Species (a non-numeric variable) as predictors.

Let's first create the dummy variables that will keep track of whether the Specie is setosa (x2.1), versicolor (x2.2) or virginica (no dummy variable is needed for this)

```{r}
a<-rep(1,50)
b<-rep(0,100)
x2.1<-c(a,b)
x2.1
d<-rep(0,50)
x2.2<-c(d,a,d)
x2.2

```

We need to create a new data frame containing all the data we need in the model.

```{r}
df <- data.frame(iris$Petal.Length,iris$Petal.Width,x2.1,x2.2)
head(df)
```


We can then fit the model as follows:

```{r}
parallel <-lm(iris.Petal.Length~iris.Petal.Width+x2.1+x2.2,data=df)
summary(parallel)
```
The estimate of the intercept for the baseline category (virginica) is the 1st coefficient of 3.4881 which is the expected Petal.Length of a virginica with 0 Petal.Width which does not make sense or not meaningful. 

The coefficient of Petall.Width of 1.0187 says a 1 cm increase in Petal.Width is expected to result in an increase of about 1.02 cm in Petal.Length for iris flowers of the same specie.  

The coefficient -2.2767 of the indicator variable x2.1 for setosa is the expected decrease in Petal.Length in cm of a setosa flower from the Petal.Length of a virginica flower of the same Petal.Width.

The coefficient -0.5789 of the indicator variable x2.2 for versicolor is the expected decrease in Petal.Length in cm of a versicolor flower from the Petal.Length of a virginica flower of the same Petal.Width.
