---
title: "Data Importing and Tidy Data"
output:
  html_document:
    df_print: paged
---
Let's import the UntidyNoHeading.csv data into R.
```{r}
df <- read.csv("UntidyNoHeading.csv")
head(df)

```

Notice that the first row is treated as the header.  To fix this, we do as follows:
```{r}
df <- read.csv("UntidyNoHeading.csv",header=FALSE)
head(df)
```
R will assign a name to each column when the data frame has no header. To rename a column, we can do as follows:

```{r}
library(dplyr)
df <- df %>% rename(Name = V1, Exam1 = V2, Exam2 = V3, Exam3 = V4)
head(df)
```
Notice that this data frame is not in tidy format because there are more columns than what is needed.We only need 3 columns for each of the variables (Name, ExamNumber and ExamScore).

To convert this untidy or messy data into a tidy data format, we do as follows:

```{r}
library(tidyr)
df_tidy <- pivot_longer(df, names_to = "Exam",values_to="Score",cols = Exam1:Exam3)
head(df_tidy)
```
Alternatively, we can also specify the columns where the values come from as follows:

```{r}
df_tidy <- pivot_longer(df, names_to = "Exam",values_to="Score",cols = c(2,3,4))
head(df_tidy)
```
We can also use negative indexing:

```{r}
df_tidy <- pivot_longer(df, names_to = "Exam",values_to="Score",cols = -1)
head(df_tidy)
```


If  you are reading ModernDive, there are more examples of data importing and converting from untidy to tidy data format in Chapter 4. There's also a discussion there of the tidyverse package which contains most of the packages we have discussed so far such as the ggplot2, dplyr, tidyr, etc.  Read this section. 