---
title: "Functions in R"
output: html_notebook
---
R has many built-in functions such as the mean(), median(), sd(), quantile(), etc
```{r}
x <- iris$Petal.Length
mean(x)
```

We can create user-defined functions in R using the function command.

```{r}
average <- function(x) {
    n <- length(x)
    val <- sum(x) / n
    return(val)
}
```
To check that the new function average() is working, let's try it with the iris$Petal.Length data.
```{r}
average(iris$Petal.Length)
mean(iris$Petal.Length)
```
As another example, let's create a new function named zscore that will compute the zscores of the elements of a numeric input y.
```{r}
zscore <- function(y) {
    out <- (y - mean(y)) / sd(y)
    return(out)
}
```

To test this function, we can again use the iris$Petal.Length as the input to this function.
```{r}
zscore(iris$Petal.Length)
```
A function can have more than one input. As an example, let's create a function that will display the smaller mean of 2 vectors x and y.

```{r}
Min_mean <- function(x, y) {
    x1 <- mean(x)
    x2 <- mean(y)
    if (x1 < x2) {
        return(list(x1, "first input has the smaller mean"))
    } else {
        return(list(x2, "2nd input has the smaller mean"))
    }
}
```

To test this function, we will call this function using the iris$Petal.Length and iris$Petal.Width as input.

```{r}
Min_mean(iris$Petal.Length, iris$Petal.Width)
```
As an exercise, create a function that will convert measurements from cm to inches or vice-versa depending on the unit of measurement of the data.  Name the function convert with input x and unit (a non-numeric object).  Test this function using the iris$Petal.Length data which are in "cm".  

Here's the R chunk that will create this function:
```{r}
convert <- function(x, unit) {
    if (unit == "cm") {
        out <- x / 2.54
        return(list(out, "inches"))
    } else if (unit == "inches") {
        out <- x * 2.54
        return(list(out, "cm"))
    }
}
```

This will test the function: 
```{r}
convert(iris$Petal.Length, "cm")
```
